Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-01-04 output/paper/manuscript.qmd
Finished first draft manuscript w/o revision. Still nothing about sensitivity to period.

Our results are consistent with prior research that associated exposure to certain \acr{edcs} with higher levels of urinary and hair cortisol [@KimLeeMoon:2018; @SearsLiuLanphear:2023]. Contrary to these prior studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and \acr{opps}, in relation to urinary glucocorticosteroids in childhood.
We are also not aware of prior studies specifically investigating the effects of exposure to \acr{edcs} and elevated levels of glucocorticosteroids in relation to \acr{hitrtse}.

[] EDCs to HITRTSE
[] EDCs to CORT.
[] CORT. to HITRTSE

---

@@

In conclusion, in a study of...

>>>>

We are not aware of prior studies specifically investigating the effects of exposure to \acr{edcs} in relation to \acr{hitrtse}. The literature on \acr{edcs} and neurodevelopment in children has mostly focused on \acr{oops} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and \acr{bpa} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to \acr{adhd} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of \acr{edcs} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with \acr{adhd} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the negative effects of several \acr{edcs} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these negative effects are stronger in males. A major limitation of these studies is the reliance on spot urine samples, that might not be representative of long-term exposures.

Our results are consistent with prior research that associated exposure to certain \acr{edcs} with higher levels of cortisol [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. There are some differences, though. First, these studies only focus on phthalates, either as individual metabolites or as mixture. Second, exposure assessment in @SunLiJin:2018 and @KimLeeMoon:2018 was performed during gestation or the first 15 months of life, respectively. Finally, the glucocorticosteroids were measured in cord blood [@SunLiJin:2018] and hair [@SearsLiuLanphear:2023]. Contrary to these studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and \acr{opps}, in relation to urinary glucocorticosteroids in childhood.

We are also not aware of prior studies specifically investigating the effects of elevated levels of glucocorticosteroids in relation to \acr{hitrtse}, although there is evidence that under- or over-production of glucocorticosteroids interfere with the normal development of the brain [@LupienMcEwenGunnar:2009]. While we did find sex-specific evidence of an effect, their clinical relevance is questionable.

@@

In conclusion, in a study of almost 1,000 children from 6 European birth cohorts, we observed that (i) exposure to non-persistent \acr{edcs} in childhood might have short-term effects on \acr{hitrtse} in childhood, (ii) exposure to non-persistent \acr{edcs} in childhood might disrupt the \acr{hpa} axis in childhood, and (iii) disruption of the \acr{hpa} axis in childhood might have short-term, sex-specific effects on \acr{hitrtse}. Future studies should investigate how glucocorticosteroids might mediate the adverse effects of exposure to \acr{edcs} on childhood neurodevelopment in larger populations.
